EGGs are adhesive!

### 2025-03-11 calco/main.tex
diagrammi

@@

What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach to the category at hand. However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, hence in $\mathcal{M}$,
while the second component does not necessarily belong to $\mathcal{M}$. 
%
Consider e.g. an EGG rule such $x \div x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}, 
can be modelled as the DPO rule below,
which can be applied with a match that is injective on equivalence classes to the left-most and the middle EGG
in the picture above
\[\mbox{to draw DPO rule, x in the same equivalence class} \]
The right-hand side is a regular mono, though, and if the starting term graph is acyclic, so is also the arriving one.
Such an asymmetry between and left- and right-hand side falls in the research about left-linear rules for 
adhesive categories that has 
been pursued e.g. in~\cite{BaldanC0G24},

>>>>

\begin{example}
		\[
                \xymatrix@C=10pt@R=6pt{
                &\emptyset \ar@{>->}[dd]|\hole_(.65){a}\ar@{>->}[rr]^{v} \ar@{>->}[dl]_{u} &&
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                                \node[style=node](v) at (0, 0){};
                                \node[style=none] at(0, 0.5){$c$};
		\end{pgfonlayer}
                \begin{pgfonlayer}{eqlayer}\draw[dashed, rounded corners] (-0.3, -0.3) rectangle (0.3, 0.3);\end{pgfonlayer}
                \end{tikzpicture} \ar@{>->}[dd]^{b} \ar@{>->}[dl]_{y} \\
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(0, 0){};
                        \node[style=none]at(0, 0.5){$b$};
                \end{pgfonlayer}
                \begin{pgfonlayer}{eqlayer}\draw[dashed, rounded corners] (-0.3, -0.3) rectangle (0.3, 0.3);\end{pgfonlayer}
        \end{tikzpicture}
                \ar@{>->}[dd]_{c}\ar@{>->}[rr]^(.7){z} & &
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(-0.5, 0){};
                        \node[style=none]at(-0.5, 0.5){$b$};
                        \node[style=node]at(0.5, 0){};
                        \node[style=none]at(0.5, 0.5){$c$};
                \end{pgfonlayer}\begin{pgfonlayer}{eqlayer}
                        \draw[dashed, rounded corners](-0.8, 0.3) rectangle (0.8, -0.3);
                \end{pgfonlayer}\end{tikzpicture}
                \ar@<2ex>@{>->}[dd]_(.3){d}\\&
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(0, 0){};
                        \node[style=none]at(0, 0.5){$a$};
                \end{pgfonlayer}
                \begin{pgfonlayer}{eqlayer}\draw[dashed, rounded corners] (-0.3, -0.3) rectangle (0.3, 0.3);\end{pgfonlayer}
        \end{tikzpicture}
                \ar@{>->}[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && 
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(-0.5, 0){};
                        \node[style=none]at(-0.5, 0.5){$a$};
                        \node[style=node]at(0.5, 0){};
                        \node[style=none]at(0.5, 0.5){$c$};
                \end{pgfonlayer}\begin{pgfonlayer}{eqlayer}
                        \draw[dashed, rounded corners](-0.8, 0.3) rectangle (0.8, -0.3);
                \end{pgfonlayer}\end{tikzpicture}
                \ar@{>->}[dl]^{n} \\
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(-0.5, 0){};
                        \node[style=none]at(-0.5, 0.5){$a$};
                        \node[style=node]at(0.5, 0){};
                        \node[style=none]at(0.5, 0.5){$b$};
                \end{pgfonlayer}\begin{pgfonlayer}{eqlayer}
                        \draw[dashed, rounded corners](-0.8, 0.3) rectangle (0.8, -0.3);
                \end{pgfonlayer}\end{tikzpicture}
                \ar@{>->}[rr]_{g} & & 
                \begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
                        \node[style=node](v)at(0, 0){};
                        \node[style=none]at(0, 0.5){$a$};
                        \node[style=node]at(1, 0){};
                        \node[style=none]at(1, 0.5){$b$};
                        \node[style=node]at(2, 0){};
                        \node[style=none]at(2, 0.5){$c$};
                \end{pgfonlayer}\begin{pgfonlayer}{eqlayer}
                        \draw[dashed, rounded corners](-.3, 0.3) rectangle (2.3, -0.3);
                \end{pgfonlayer}\end{tikzpicture} }
\]

@@

What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach to the category at hand. However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, hence in $\mathcal{M}$,
while the second component does not necessarily belong to $\mathcal{M}$. 
%
Consider e.g. an EGG rule such $x \div x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}, 
can be modelled as the DPO rule below,
which can be applied with a match that is injective on equivalence classes to the left-most and the middle EGG
in the picture above
\[\xymatrix{
                \begin{tikzpicture}[baseline=(w)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
        \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
        \end{pgfonlayer}\end{tikzpicture}
        & \begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
	\ar[l] \ar[r] &
        \begin{tikzpicture}[baseline=(nil.center)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
                \node[style=small box] at (0, -1){$1$};
                \node[style=none](one) at (0.3, -1){};
                \node[style=node](z) at (0.9, -1){};
                \node[style=none](nil) at (0, -0.5){};
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
                \draw(one.center) to (z);
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \draw[dashed, rounded corners](0.6, 0.3) rectangle (1.2, -1.3);
        \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
        \end{pgfonlayer}\end{tikzpicture}
}\]\end{center}

The right-hand side is a regular mono, though, and if the starting term graph is acyclic, so is also the arriving one.
Such an asymmetry between and left- and right-hand side falls in the research about left-linear rules for 
adhesive categories that has 
been pursued e.g. in~\cite{BaldanC0G24},
